Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-13 aistats.tex
Small fixes in Example section

\subsection{Asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
We now review classical asymptotic results. 
We point the reader towards \citet[\S1.1]{ostrovskii2021finite} for a more comprehensive overview.
Asymptotic results typically show that the MLE expressed with natural parameters $\nat$ is asymptotically normal: it reaches the Cramér-Rao lower bound \citep[for instance Ch4.2]{vdv1998asymptotic}.
With exponential families, the MAP is more simply expressed with mean parameters $\meanp$.
We get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
\alignn{
    \bregmanconj(\m^* ; \m)
    &= \frac{\norm{\m^* - \m}^2_{\mF}}{2}
    + O(\norm{\m - \m^*}^3),
    \label{eq:bregman_asymptote}
}
where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum. It happens that  $\mF$ is also the inverse \textit{Fisher information matrix} at $\nat^*$, since
\aligns{
    \mF
    :=\nabla^2\conj(\m^*)
    = \nabla^2\logpart(\nat^*)^{-1}
    = \Cov_{\nat^*}[T(X)]^{-1}  \; .
}
Plugging~\eqref{eq:defMLE} or~\eqref{eq:defMAP} into~\eqref{eq:bregman_asymptote}, one can show that MLE and MAP verify 
\begin{align}
	\label{eq:MAP_asymptote}
	\E \bregmanconj \left (\E [T(X)] ; \hat \meanp_n^\text{MLE/MAP} \right )
	= \frac{d}{2n} + O(n^{- \frac{3}{2}}) \; .
\end{align}
See proof in the Appendix.
Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
This asymptote is also independent of the optimum $\meanp^*$.

@@

{\bf If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex \citep{kakade2009duality}), then
\begin{align}
    \bregmanconj(\m^* ; \m)
    \leq \frac{L}{2} \norm{\m^* - \m}^2
\end{align}
so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
Note that it is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
\TODO{connection with optim smooth rate.}

\subsection{Locally Quadratic Case}
\textcolor{red}{Check this! In our case, since we assume $A$ and $A^*$ to be strictly convex,}
all Bregman divergences are locally quadratic.
Under some assumptions, like self-concordance of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in.
\begin{proposition}
Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then%\footnote{$0.21$ is the value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}
\aligns{
	\norm{\m^*-\m}_{\mF} < 0.21 
	\implies
	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2
}
\end{proposition}
Proof in the appendix.
We know that $\conj$ is self-concordant when $T$ is 1-dimensional and $\logpart$ is self-concordant.
This is true for
exponential distributions,
Gaussians with known mean,
Laplace with known mean,
Pareto with known minimum value,
or Weibull with known shape $k$. 
The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic}.

@@

\section{Examples}
\label{sec:example}

\subsection{Gaussian Variance}
A core example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
Its density reads 
\[
	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}.
\]
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
\textcolor{red}{The explanation here is not clear enough, or too concise} Now we can match the log-likelihood with the exponential family template to get the log-partition function, and taking the conjugate to find the entropy
\begin{align}
	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi), \\
	\conj(\m) &= \half\left( -\log(\m) + \log\frac{\pi}{2} - 1 \right) \; .
\end{align}
Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative Gamma distribution.
It also means $\bregman$ and $\bregmanconj$ have the same shape
\begin{align}
	\bregmanconj( \m_*; \m_n)
	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right), \\
	\bregman( \nat_n; \nat_* )
	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
Intuitively, the divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
\begin{align}
	\phi(z) := \half (z - 1 - \log(z))
\end{align}
illustrated in Figure~\ref{fig:phi}.

@@

Theorem \ref{eq:MLE_rate} reports upper bounds on the expected value of this divergence for the MLE and the MAP.

\begin{theorem}[MLE Tight Bound]
	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}

This upper bound is asymptotically tight.
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
We get a similar bound for the multivariate generalization :
the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.

For the MAP, we get an upper bound thanks to the inequality $ - \log(z) \leq \inv{z} - 1$, which implies
\begin{align}
	\label{eq:log_bound}
	\phi(z) \leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z} \; .
\end{align}

\begin{theorem}[MAP Bound]
For $n\in \naturalnumbers$, the expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
 \begin{align}
	& \expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}\\
	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
\end{align}
\end{theorem}

This inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.  For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$. This correlates well with numerical observations.
Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$, therefore losing a factor 2 compared to $1/2n$.

>>>>

\subsection{Asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
We now review classical asymptotic results. 
We point the reader towards \citet[\S1.1]{ostrovskii2021finite} for a more comprehensive overview.
Asymptotic results typically show that the MLE expressed with natural parameters $\nat$ is asymptotically normal: it reaches the Cramér-Rao lower bound \citep[for instance Ch4.2]{vdv1998asymptotic}.
With exponential families, the MAP is more simply expressed with mean parameters $\meanp$.
We get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
\alignn{
    \bregmanconj(\m^* ; \m)
    &= \frac{\norm{\m^* - \m}^2_{\mF}}{2}
    + O(\norm{\m - \m^*}^3),
    \label{eq:bregmanTaylor}
}
where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum. It happens that  $\mF$ is also the inverse \textit{Fisher information matrix} at $\nat^*$, since
\aligns{
    \mF
    :=\nabla^2\conj(\m^*)
    = \nabla^2\logpart(\nat^*)^{-1}
    = \Cov_{\nat^*}[T(X)]^{-1}  \; .
}
Plugging~\eqref{eq:defMLE} or~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, one can show that MLE and MAP verify 
\begin{align}
	\label{eq:MAP_asymptote}
	\E \bregmanconj \left (\E [T(X)] ; \hat \meanp_n^\text{MLE/MAP} \right )
	= \frac{d}{2n} + O(n^{- \frac{3}{2}}) \; .
\end{align}
See proof in the Appendix.
Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
This asymptote is also independent of the optimum $\meanp^*$.

@@

{\bf If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex \citep{kakade2009duality}), then
\begin{align}
    \bregmanconj(\m^* ; \m)
    \leq \frac{L}{2} \norm{\m^* - \m}^2
\end{align}
so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
Besides the Gaussian with known mean, no standard exponential family verify this assumption either.
Note that it is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
\TODO{connection with optim smooth rate.}

\subsection{Locally Quadratic Case}
FromTaylor expansion~~\eqref{eq:bregmanTaylor}, 
we know that all Bregman divergences are locally quadratic.
Under some assumptions, like self-concordance of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in.
\begin{proposition}
Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then
\aligns{
	\norm{\m^*-\m}_{\mF} < 0.21 
	\implies
	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2
}
\end{proposition}
Proof in the appendix
\footnote{In case the reader is intrigued, $0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
We know that $\conj$ is self-concordant when $T$ is 1-dimensional and $\logpart$ is self-concordant.
This is true for
exponential distributions,
Gaussians with known mean,
Laplace with known mean,
Pareto with known minimum value,
or Weibull with known shape $k$. 
The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic}.

@@

\section{Gaussian Variance Example}
A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
Its density reads 
\[
	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}.
\]
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
\textcolor{red}{The explanation here is not clear enough, or too concise} Now we can match the log-likelihood with the exponential family template to get the log-partition function, and taking the conjugate to find the entropy
\begin{align}
	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi), \\
	\conj(\m) &= \half\left( -\log(\m) + \log\frac{\pi}{2} - 1 \right) \; .
\end{align}
Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative Gamma distribution.
It also means $\bregman$ and $\bregmanconj$ have the same shape
\begin{align}
	\bregmanconj( \m_*; \m_n)
	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right), \\
	\bregman( \nat_n; \nat_* )
	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
Intuitively, the divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
\begin{align}
	\phi(z) := \half (z - 1 - \log(z))
\end{align}
illustrated in Figure~\ref{fig:phi}.

@@

In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
\begin{theorem}[MLE Tight Bound]
\label{thm:varianceMLE}
	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
This upper bound is asymptotically tight.
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
We get a similar bound for the multivariate generalization :
the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$ (see appendix).

For MAP, we get an upper bound thanks to
\begin{align}
	\label{eq:log_bound}
	\phi(z) \leq \phi(z) + \phi(\inv{z}) =  \half (z + \inv{z}) - 1 \; .
\end{align}

\begin{theorem}[MAP Bound]
\label{thm:varianceMAP}
For $n\in \naturalnumbers$, the expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
 \begin{align}
	& \expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}\\
	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
\end{align}
\end{theorem}
\rlp{upper bound and compare the symmetrized bregman instead?}
This inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.  For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$. This correlates well with numerical observations (see appendix).
Remark that in using~\eqref{eq:log_bound} we lost a factor $2$ and got a rate$\inv{n}$ instead of $\inv{2n}$.

These results are instantiation the bias-variance decomposition and giving us a flavor of what we would like as a general result for exponential families : a finit sample convergence rate, with a variance and a bias terms that reflect the important constants of the problem.
In the next section, we investigate what this problem is brining to optimization.
